---
title: "2.inspect_data"
author: "Dennis Kim"
date: "2022-10-03"
output: html_document
---

# Objective 

To manipulate the merged tracking data with the other addtional columns for SSFs. To do this, I will: 

1. Determine infected individuals and healthy individuals    
3=2. Inspect movement characteristics of individuals 

# Document Preamble 
```{r preamble, message=FALSE, warning=FALSE}
# load libraries
library(knitr)
library(dplyr)
library(readr)
library(data.table)
library(DT)
library(here)
library(stringr)
library(tidyr)
library(purrr)
library(lubridate)
library(amt)
library(ggplot2)
library(raster)
library(sf)
options(width = 150)

# set knitr options 
opts_chunk$set(fig.width = 6, fig.height = 5, comment = NA)
```

# Data download and initial preparation

Read in the gps data and 
```{r read in data}
# Location data of the oryx with additional disease columns 
oryx <- read.csv(here::here("data", "oryx_gps_id.csv"))

# NAs in fate column convert to unkown
oryx$fate[is.na(oryx$fate)] <- "unknown"

# summary of the location data 
summary(oryx)

oryx %>%  filter(Rgroup == 4) %>% filter(cause_of_death %in% c("disease", "presumed disease")) %>% distinct(masterID)
```

# Multiple Individuals 

Referenced by Katherine Mertes -- select healthy and infected individuals within the same release group (R4)
```{r indicators}
# check the disease individuals -- that are dead - filter individuals that have at least 1000 location points
disease_id <- oryx %>% filter(Rgroup == 4) %>% filter(cause_of_death %in% c("disease", "presumed disease")) %>% nest_by(masterID) %>% mutate(rows = nrow(data)) %>% dplyr::select(-data) #%>% filter(rows > 1000) 

# same for the susceptible individuals 
susceptible_id <- oryx %>% filter(Rgroup == 4) %>% filter(cause_of_death %in% "alive") %>% nest_by(masterID) %>% mutate(rows = nrow(data)) %>% dplyr::select(-data) #%>% filter(rows > 10000)

# check if the body condition data are available
## read the body condition data 
body_condition <- read.csv(here::here("data", "oryx_individual_field_data_122322.csv"))

# only select the columns that are applicable to the data 
body_condition <- body_condition %>% dplyr::select(datetime, animalID, lat, long, body_condition)

# filter the data with non-NAs for body condition values 
body_condition <- body_condition[complete.cases(body_condition[, 5]),]

# check the available body condition individuals based on Rgroup 
body_condition_id <- body_condition %>% nest_by(animalID) %>% mutate(rows = nrow(data)) %>% dplyr::select(-data) %>% 
  mutate(group = case_when(
    animalID %in% disease_id$masterID ~ "Infected",
    animalID %in% susceptible_id$masterID ~ "Susceptible"
  )) %>% 
  na.omit()

body_condition_id %>% filter(group == "Infected")

# save the body_condition id 
#readr::write_rds(body_condition_id, path = here("data/finalized_data", "bc_id.Rdata"))
```


```{r select individuals}
# select R4 group that has the body condition data (30 individuals)
multiple_oryx <-
  oryx %>% filter(
    masterID %in% body_condition_id$animalID
  )

multiple_oryx

# remove the original data to reduce the memory capacity 
#rm(oryx)
#rm(disease_id)
#rm(susceptible_id)
#rm(body_condition_id)

# change the time format of the data 
multiple_oryx$date <- multiple_oryx$date <- as.POSIXct(multiple_oryx$date, format = "%Y-%m-%d %H:%M:%S")
multiple_oryx$date_collared <- multiple_oryx$date_collared %>% mdy()
multiple_oryx$date_released <- multiple_oryx$date_released %>% mdy()

# get rid of any missing location values in long, lat, and timestamp 
filter_ids <- multiple_oryx %>% dplyr::select("long", "lat", "date") %>% complete.cases
multiple_oryx <- multiple_oryx %>% filter(., filter_ids == TRUE)

multiple_oryx %>% distinct(cause_of_death)

# create id group columns 
multiple_oryx <- multiple_oryx %>% mutate(Igroup = ifelse(
  cause_of_death %in% "alive",
  "susceptible",
  "infected"
))

multiple_oryx
```

Add shrubs and trees layers 
```{r shrub and trees}
# call the files as a vector of raster files 
#imgs <- list.files("E:/disease_oryx/data/env_data/shrub", "tif$", full.names = TRUE)

# use lapply to create raster objects wrapped in terra::src to create a SpatRasterCollection
#ic <- sprc(lapply(imgs, rast))

# pass the image collection to the mosaic function
#shrub_raster <- terra::mosaic(ic)
#shrub_raster <- raster(shrub_raster) # convert the SpatRasterlayer class to RasterLayer

# save the shrub ratster 
#writeRaster(shrub_raster, filename = "E:/disease_oryx/data/env_data/shrub/final.shrub.tif", format = "GTiff")

# read the saved merged final raster layer 
shrub_raster <- raster("D:/Oryx/data/env_data/shrub/final.shrub.tif")
crs(shrub_raster)
plot(shrub_raster)
# reproject 
#shrub_raster1 <- projectRaster(shrub_raster, crs = "+proj=longlat +datum=WGS84")
#shrub_raster1

#writeRaster(shrub_raster1, filename = "E:/disease_oryx/data/env_data/shrub/final.shrub1.tif", format = "GTiff")

# plot the shrub raster
#plot(shrub_raster1)

# extract the shrub layers 
#multiple_oryx <- cbind(extract(shrub_raster, multiple_oryx[,c("utm_e", "utm_n")], df = T), multiple_oryx) %>% dplyr::select(-ID)

#multiple_oryx
```

# Create tracks in amt

Add a track class to the data and summarize the data 
```{r amt conversion}
# make tracks 
trk <- multiple_oryx %>% 
  amt::make_track(
    .,
    .x = utm_e,
    .y = utm_n,
    .t = date,
    masterID = masterID,
    collar_id = collar_ID,
    sex = sex,
    fate = fate,
    date_collared = date_collared,
    date_released = date_released,
    temp_c = temp_c,
    prev_collar_ID = prev_collar_ID,
    date_released = date_released,
    cause_of_death = cause_of_death,
    Igroup = Igroup,
    crs = st_crs(4326)
  ) 

trk

# save the tracking data 
#write_rds(trk, path = here("data","multiple_original_trk.Rdata"))

# check the sampling rates 
trk %>% nest(data = -"masterID") %>%  mutate(sr = map(data, summarize_sampling_rate)) %>%
  dplyr::select(masterID, sr) %>%
  unnest(cols = sr) %>% print(n = nrow(trk))

# calculate the step lengths and speed to see if it's reasonable 
tr_check <-
  trk %>% nest(data = -"masterID") %>% 
  mutate(data = map(data, ~ .x %>% steps())) %>%
  unnest(cols = data) %>%
  mutate(speed = ((sl_) / as.numeric(dt_)))

tr_check %>% head()

##check the summary of the data
summary(tr_check$speed) ##Check the spread of the speed
summary(tr_check$sl_)
summary(as.numeric(tr_check$dt_))

# speed distribution
ggplot(tr_check) + geom_histogram(aes(x = speed), fill = "blue") + theme_bw()
```

# Resample the tracking data and generate the random points 
```{r sampling rate}
# check the sampling rate of individuals - 1 hour 
trk %>% amt::summarize_sampling_rate_many(c("masterID"))

# regularize the steps and generate random steps 
ssfdat <- trk %>% nest(data = -masterID) %>% 
  mutate(
    data = map(
      data,
      ~ .x %>% 
        track_resample(rate = hours(1), tolerance = minutes(30)) %>%
        filter_min_n_burst(min_n = 3) %>%
        steps_by_burst() %>%
        random_steps() %>% 
        extract_covariates(shrub_raster)
      )
    ) %>% unnest(cols = data) 

ssfdat <- ssfdat %>% mutate(
  log_sl_ = log(sl_),
  cos_ta_ = cos(ta_),
  speed = sl_ / (as.numeric(dt_))
)

glimpse(ssfdat)
```

# seasons 
```{r seasons}
# create an indicator column that label the different seasonal ranges
ssfdat1 <- ssfdat %>%
  mutate(
    week = week(t1_),
    day_of_year = yday(t1_),
    season_level = case_when(
      dplyr::between(day_of_year, 72, 191) ~ 'hotdry',
      dplyr::between(day_of_year, 192, 274) ~
        'rainy'
    )
  ) %>%
  mutate(season_level = replace_na(season_level, "cooldry")) %>%
  ungroup()

ssfdat1 %>% glimpse()
```

# body condition 

filter the body condition data for the selected individuals and join the data 
```{r R01M Bc}
# change the column name of the body condition same as the original tracking data 
colnames(body_condition) <- c("t1_", "masterID", "y1_", "x1_", "body_condition")

# select the healthy individuals for our cases 
BC <-
  body_condition %>% filter(
    masterID %in% body_condition_id$animalID
  )

# create julian dates and week id for the BC data
BC <- BC %>%
  mutate(julian = yday(t1_),
         week = week(t1_)) 

# calculate the average body condition value based on each duplicated week
average.week.BC <- BC %>% group_by(masterID, week) %>% 
  mutate(avg.bc = mean(body_condition)) %>% 
  dplyr::select(week, avg.bc) %>% 
  distinct()

average.week.BC

# join the data
ssfdat2 <- ssfdat1 %>% left_join(average.week.BC, by = c("masterID", "week"))
ssfdat2 %>% na.omit(.)

# only select complete data with body condition
comp.trk <- ssfdat2 %>% filter(!is.na(avg.bc)) %>% 
  mutate(c_avg_BC = ifelse(avg.bc < 4, "under", ifelse(avg.bc < 7, "optimal", "over")) %>% as.factor(.))

comp.trk
```

# Individual selection for further analysis
Save the selected individuals' track data
```{r save the files}
#write_rds(comp.trk, path = here("data/finalized_data","multiple_trk.Rdata"))
```

## Footer
```{r footer}
sessionInfo()
```