---
title: "3.identify_exp_season"
author: "Dennis Kim"
date: "2022-10-03"
output: html_document
---

# Objective 

To manipulate the merged tracking data with the other addtional columns for SSFs. To do this, I will: 

1. Determine infected individuals and healthy individuals    
2. Identify different levels of experiences among each groups and different seasonality  
3. Select individuals that are applicable for further analysis 

# Document Preamble 
```{r preamble, message=FALSE, warning=FALSE}
# load libraries
library(knitr)
library(dplyr)
library(readr)
library(data.table)
library(DT)
library(here)
library(stringr)
library(tidyr)
library(purrr)
library(lubridate)
library(amt)
library(ggplot2)
library(RColorBrewer)
library(mcp)
options(width = 150)

# set knitr options 
opts_chunk$set(fig.width = 6, fig.height = 5, comment = NA)
```

# Initial data preparation

Read in trk data and select the healthy individual: 21826.2 
```{r read in data}
# Location data of the oryx with additional disease columns 
oryx.healthy <- read_rds(here("data", "trk.healthy.Rdata"))
oryx.healthy

# select the individual: 21826.2
#healthy.id <- oryx.healthy %>% filter(id == 21826.2)

# summary of the selected individual 
#healthy.id %>% dplyr::select(date_released) %>% distinct()
```

# Segment movements based on the different released time periods 

*[note]*: individuals who were released in 2016-2017, then darted in the field and re-collared in 2021-2022 - we need to segment movements based on the time periods
- "ranging": released in August 2016 (home ranging formed: ~27 weeks)
- "penned": released in Jan 2017 (home ranging formed: ~ 35 weeks)
- "experienced": later released (2021-2022)

Visualize the approx. time periods of the collaring: all the selected healthy individuals are released in 2016 - thus, there is no "penned" individuals in the current dataset.
```{r healthy mv vis}
# sl distribution over time 
ggplot()+
  geom_density(data = oryx.healthy, aes(as.Date(t_), fill = sl_))+
  facet_wrap(~id, scales = "free_x")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  scale_x_date(date_labels = "%y", date_breaks = "1 year")
```
Group the data by experience groups
```{r experience group}
# ranging group: less than 2019
oryx.healthy1 <- oryx.healthy %>% mutate(group = ifelse(year < 2019, "ranging", "experienced"))

oryx.healthy1 %>% glimpse()

# visualize the group of movements 
ggplot(oryx.healthy1,aes(x_, y_, color = group, group = group))+
  geom_point()+
  facet_wrap(~id)
```

# Group by the seasonal ranges 

Seasonal ranges 
- Hot dry season: 03.13 ~ 07.10 (day of year 72-191)
- Rainy season: 07.11 ~ 10.1 (day of year 192-274)
- Cool dry season: 10.02 ~ 03.12 (day of year 275-71)

In our current dataset, only the rainy and cooldry seasons are present. 4 individuals were tracked during the rainy seasons while 1 individual was collared during the cool dry season. 
```{r seasonal ranges}
# create an indicator column that label the different seasonal ranges
oryx.healthy2 <- oryx.healthy1 %>% 
  mutate(day_of_year = yday(date_released)) %>% 
  mutate(season_level = case_when(dplyr::between(day_of_year, 72, 191)~'hotdry',
                                  dplyr::between(day_of_year, 192, 274)~'rainy')) %>% 
  mutate(season_level = replace_na(season_level, "cooldry"))

# check the unique seasons across individuals: in our cases, only rainy and cooldry seasons are available 
oryx.healthy2 %>% dplyr::select(season_level) %>% distinct()

# visualize the group of movements 
ggplot(oryx.healthy2, aes(x_, y_, color = season_level, group = season_level))+
  geom_point()+
  scale_color_brewer(palette = "Dark2")+
  facet_wrap(~id)
```

save the grouped data for further analysis 
```{r save the files}
#write_rds(oryx.healthy2, path = here("data","trk.healthy.filtered.Rdata"))
```


## Footer
```{r footer}
sessionInfo()
```