---
title: "Unconstrained HHMM"
author: "Dennis Kim"
date: "2025-01-16"
output: html_document
---

# Document Preamble 
```{r preamble, message=FALSE, warning=FALSE}
# load libraries
library(knitr)
library(dplyr)
library(readr)
library(here)
library(tidyr)
library(momentuHMM)
library(mitools)
library(data.tree)
library(DiagrammeR)

options(width = 150)
```

# 1. Format of tracking data

prepare data for HMM format
```{r HMM formats}
data_hhmm <- readRDS(here("data", "04.11.2025_data_HHMM.Rdata"))
```

# 2. HHMM structure  

Our case has no data streams observed at the coarse-scale level (infected vs non-infected).

## 2.1. Hierarchical data stream structure

Specify the probability distribution for the fine-scale step length and turn angle data stream via the hierDist data tree Node 
```{r data stream structure}
# data stream distributions: 
## level 1 = coarse level (no data streams)
## level 2 = fine level (step="gamma", angle = "vm")

hierDist <- data.tree::Node$new("oryx disease HHMM dist")
hierDist$AddChild(name="level1")
hierDist$AddChild(name="level2")
hierDist$level2$AddChild(name="step", dist = "gamma")
hierDist$level2$AddChild(name = "angle", dist = "vm")

print(hierDist, "dist")
#plot(hierDist)
```

## 2.2. Hierarchical state structure

define HHMM strcture via the hierStates data tree Node
```{r HHMM structure}
# define hierarchical HMM: 
## state 1-2 = coarse state 1 (non-infected)
## state 3 = corase state 2 (infected)
## state 4 = coarse state 3 (death) 

hierStates <- data.tree::Node$new("oryx disease HHMM states")
hierStates$AddChild(name="nonInfected")
hierStates$nonInfected$AddChild(name="e", state=1) # exploring
hierStates$nonInfected$AddChild(name="r", state=2) # resting
hierStates$AddChild(name="Infected")
hierStates$Infected$AddChild(name="i", state=3) # infected
hierStates$Infected$AddChild(name="id", state = 4) # infected dummy
hierStates$AddChild(name="Death")
hierStates$Death$AddChild(name="d", state = 5) # death
hierStates$Death$AddChild(name="dd", state =6) # death dummy

print(hierStates, "state")
#plot(hierStates)
```

## 2.3. Data stream probability distribution

We assume the fine-scale data stream probability distribution depends on coarse-scale state. 

### 2.3.1. Setting up the starting values 
```{r starting values}
# Number of states (leaves in the hierarchical tree)
nbStates <- length(hierStates$Get("states", filterFun = data.tree::isLeaf))
nbStates

# Defining starting values
# Initialize start values for step length (gamma) and angle (vm) distributions
# 6 fine-scale states: e, r, i, id, d, dd

# Step length parameters
mu0 <- c(1800, 380, 25, 25, 4, 4)  # step mu for e, r, i/id, d/dd
sd0 <- c(1050, 370, 20, 20, 4, 4)  # step sd for e, r, i/id, d/dd

# Turn angle parameters
amu0 <- c(-0.03, 0.03, -3, -3, 3, 3)  # angle mean for e, r, i/id, d/dd
kappa0 <- c(1.5, 0.16, 0.36, 0.36, 0.3, 0.3)  # angle concentration for e, r, i/id, d/dd

Par0 <- list(
  step = c(mu0, sd0), 
  angle = c(amu0, kappa0)
)

# Check the structure
str(Par0)
```

constrain fine-scale data stream distributions to be same using DM
```{r DM}
# DM - constrain fine-scale data stream distributions to be same 
# so that infected (i) and infected dummy (id) share the same parameters,
# and death (d) and death dummy (dd) share the same parameters.

#####################################################
## Pseudo-design matrix for observation parameters ##
#####################################################
# Step length
dw_DM_step <- matrix(0, nrow = 2*nbStates, ncol = 2*(nbStates - 2))
dw_DM_step[1,1] <- 1 # mu1
dw_DM_step[2,2] <- 1 # mu2
dw_DM_step[3:4,3] <- 1 # mu3 = mu4
dw_DM_step[5:6,4] <- 1 # mu5 = mu6
dw_DM_step[7,5] <- 1 # sd1
dw_DM_step[8,6] <- 1 # sd2
dw_DM_step[9:10,7] <- 1 # sd3 = sd4
dw_DM_step[11:12,8] <- 1 # sd5 = sd6
rownames(dw_DM_step) <- paste0(rep(c("mu", "sd"), each = nbStates), 1:nbStates)
colnames(dw_DM_step) <- paste0(rep(c("mu", "sd"), each = 4),
                               c("1", "2", "34", "56"), ":(Intercept)")

# Turning angle (constraints are the same as step length)
# Turning angle (von Mises distribution)
dw_DM_angle <- matrix(0, nrow = 2*nbStates, ncol = 2*(nbStates - 2))
dw_DM_angle[1,1] <- 1 # amu1
dw_DM_angle[2,2] <- 1 # amu2
dw_DM_angle[3:4,3] <- 1 # amu3 = amu4
dw_DM_angle[5:6,4] <- 1 # amu5 = amu6
dw_DM_angle[7,5] <- 1 # kappa1
dw_DM_angle[8,6] <- 1 # kappa2
dw_DM_angle[9:10,7] <- 1 # kappa3 = kappa4
dw_DM_angle[11:12,8] <- 1 # kappa5 = kappa6

rownames(dw_DM_angle) <- paste0(rep(c("amu", "kappa"), each = nbStates), 1:nbStates)
colnames(dw_DM_angle) <- paste0(rep(c("amu", "kappa"), each = 4),
                                c("1", "2", "34", "56"), ":(Intercept)")

# Define DM list for each data stream, applying the constraint
DM <- list(
  step = dw_DM_step,
  angle = dw_DM_angle
)

# Get initial parameter values for data stream probability distributions
Par <- getParDM(
  data_hhmm, 
  hierStates = hierStates, 
  hierDist = hierDist, 
  estAngleMean = list(angle = TRUE),
  Par = Par0, 
  DM = DM,
)

Par
```

### 2.3.2. Covariates

include shrub cover and time since release covariates on the fine-scale state transiiton probabilities - specify this via the hierFormula argument:
```{r covariates}
# define hierarchical t.p.m. formula(s)
hierFormula <- data.tree::Node$new("oryx disease HHMM formula")
hierFormula$AddChild(name = "level1")
hierFormula$AddChild(name= "level2", formula = ~ shrub + time_since_release)
```

### 2.3.3. Transitional probability matrix (t.p.m)

#### 2.3.3.1. Set initial parameter values
Assume transition probabilities between coarse-scale states are distinct. Set initial parameter values. Define separate transition probability matrices for coarse-scale and fine-scale states via hierBeta argument:
```{r initial parameter tpm}
# Define the transition matrix for the coarse-level states (Non-Infected, Infected, Death)
hierBeta <- data.tree::Node$new("oryx beta")

# Add coarse-level state transition matrix
hierBeta$AddChild(name = "level1",
                  beta = matrix(c(
                      -2,  # s > i
                      -5, # s > d 
                      -2, # i > s
                      -3, # i > d
                      -1e6, # d > s
                      -1e6), # d> i
                      nrow = 1, 
                      ncol = 6, 
                      byrow = TRUE))

# fine scale level
#Exploring State (1):
##Can transition to the Resting state (2) or remain in Exploring (1).

# Resting State (2):
##Can transition to the Exploring state (1) or reamin in Resting (2).

# Infected State (3):
##Individuals cannot leave this state.

# Death State (4)
##Individuals cannot leave this state

hierBeta$AddChild(name="level2")
# Fine-scale transition matrix for Non-Infected (with covariates)
hierBeta$level2$AddChild("nonInfected",
                         beta = matrix(c(
                             # Transition from Exploring -> Resting (covariates: shrub, time_since_release)
                             -2, 0, 0, 
                             # Transition from Resting -> Resting (covariates: shrub, time_since_release)
                             -2, 0, 0
                         ), nrow = 3, ncol = 2))

# Fine-scale transition matrix for Infected (with covariates)
hierBeta$level2$AddChild("Infected",
                         beta = matrix(c(
                             # Transition from Infected -> ID (covariates: shrub, time_since_release)
                             -1e6, -1e6, -1e6,
                             # Transition from ID -> ID (covariates: shrub, time_since_release)
                             -1e6, -1e6, -1e6
                         ), nrow = 3, ncol = 2))

# Fine-scale transition matrix for Death (with covariates)
hierBeta$level2$AddChild("Death",
                         beta = matrix(c(
                             # Transition from Death -> DD (covariates: shrub, time_since_release)
                             -1e6, -1e6, -1e6,
                             # Transition from DD -> DD (covariates: shrub, time_since_release)
                             -1e6, -1e6, -1e6
                         ), nrow = 3, ncol = 2))
```

#### 2.3.3.2. Define parameter constraints
```{r paramaeter constraint tpm}
fixBeta <- data.tree::Node$new("tpm constraints")
# No transition from coarse scale infected to non-infected to death
fixBeta$AddChild(name = "level1",
                 fixPar = matrix(c(NA, 
                                   NA, 
                                   NA,
                                   NA,
                                   -1e6,
                                   -1e6), nrow = 1))

fixBeta$AddChild(name = "level2")
fixBeta$level2$AddChild(
    "nonInfected",
    fixPar = matrix(c(NA, NA, NA, # Exploring > Resting
                      NA, NA, NA # Resting > Resting
                      ),
                    nrow = 3, ncol = 2))

fixBeta$level2$AddChild(
    "Infected",
    fixPar = matrix(c(-1e6, -1e6, -1e6, # Infected > Infected dummy
                      -1e6, -1e6, -1e6 # Infected dummy > Infected dummy
                      ),
                    nrow = 3, ncol = 2))

fixBeta$level2$AddChild(
    "Death",
    fixPar = matrix(c(-1e6, -1e6, -1e6, # Death > Death dummy
                      -1e6, -1e6, -1e6 # Death dummy > Death dummy
                      ),
                    nrow = 3, ncol = 2))

# Ensure fixBeta matrix is constrained.. 
fixBeta$level2$nonInfected$fixPar
```

# 2.3. Fit the HHMM

```{r check HHMM}
# check hierarchical model specification and parameters 
momentuHMM::checkPar0(data_hhmm,
                      hierStates = hierStates,
                      hierDist = hierDist,
                      hierFormula = hierFormula,
                      Par0 = Par,
                      DM = DM,
                      hierBeta = hierBeta,
                      fixPar = list(beta = fixBeta),
                      estAngleMean = list(angle=TRUE))
```

```{r fit model}
# fit hierarchical HMM 
hhmm <- momentuHMM::fitHMM(data_hhmm,
                      hierStates = hierStates,
                      hierDist = hierDist,
                      hierFormula = hierFormula,
                      Par0 = Par,
                      DM = DM,
                      hierBeta = hierBeta,
                      fixPar = list(beta = fixBeta),
                      estAngleMean = list(angle=TRUE))

# Print parameter estimates
hhmm

# save the hmm output 
readr::write_rds(hhmm, path = here("data", "05.13.2025.4s_unconstrained_HHMM.Rdata"))
```

save the sequence plot
```{r state spatial visualization}
# generate state sequence
states <- momentuHMM::viterbi(hhmm)

# check the length
length(states)
nrow(data_hhmm)

# generate the most likely state squences at each level of the hierachy 
hStates <- momentuHMM::viterbi(hhmm, hierarchical = TRUE)
lapply(hStates, length)

# include the sequence states into the dataset
data_hhmm$states <- states

unique(data_hhmm$states)

data_hhmm$states <- factor(states, levels = c("1", "2", "3", "4", "5", "6"))

data_hhmm %>% summary()

# save the rds file for the state sequences 
readr::write_rds(data_hhmm, path = here("data", "05.13.2025.4s_unconstrained_HHMM_viberti.Rdata"))
```

# Footer
```{r footer}
sessionInfo()
```