---
title: "10.Simulation_SIR"
author: "Dennis Kim"
date: "2024-01-30"
output: html_document
---

# Document Preamble 
```{r preamble, message=FALSE, warning=FALSE}
# load libraries
library(knitr)
library(readr)
library(data.table)
library(here)
library(purrr)
library(lubridate)
library(scales)
library(conflicted)
library(tidyverse)
library(scico)
library(patchwork)
library(tictoc)

# call environmental data
library(terra)
library(sf)

# analysis
library(amt)
library(MASS)
library(statip)

# visualization 
library(ggplot2)
library(tidyterra)

options(width = 150)

conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflicts_prefer(lubridate::hour)
conflicts_prefer(terra::plot)
conflicts_prefer(dplyr::lag)
```

# call the shrub layers 
```{r environmental layers}
# observed shrub cover
shrub <- terra::rast("D:/oryx_hmm/data/shrub/final.shrub.tif")
```


# simulating tracks using the amt functions 

1. specifying paamters governing the movements for the individual using the *make_issf_model()** function 
2. specifying a redistribution kernel for generating movements based on an initial location and bearing 
3. generating paths from the simulated redistribution kernel specified for a user-provided number of steps 

visualize the movement parameter distribution per each state
```{r vis movement params}
# gamma distribution

## data for plotting
plot_sl <- data.frame(x = rep(NA, 100))

## x-axis is sequence of possible step lengths
plot_sl$x <- seq(from = 0, to = 2500, length.out = 100)

# y-axis is the prob density under the given gamma distribution 

# Susceptible (S)
plot_sl$s <- dgamma(
  x = plot_sl$x,
  shape = 0.5,
  scale = 800
)

# Infected (I)
plot_sl$i <- dgamma(
  x = plot_sl$x,
  shape = 0.2,
  scale = 300
)

# Recovered (R)
plot_sl$r <- dgamma(
  x = plot_sl$x,
  shape = 0.6,
  scale = 700
)

# Dead (D)
plot_sl$d <- dgamma(
  x = plot_sl$x,
  shape = 0.1,
  scale = 50
)

# pivot from wide data to long data
plot_sl <- plot_sl %>% pivot_longer(cols = -x)
plot_sl


# plot 
ggplot(plot_sl, aes(x = x, y = value, color = factor(name)))+
  geom_line(size = 1)+
  xlab("Step Length (m)")+
  ylab("Probability Density")+
  scale_color_manual(name = "Distribution",
                     breaks = c("s", "i", "r", "d"),
                     values = c("blue", "orange", "red", "darkgreen"))+
  theme_bw()
```

prep for the simulation 
```{r prep for the simulation}
# covariance matrix for the movement parameters 
sig <- matrix(c(0.4, 0.15, 0.15, 0.15, 0.3, 0.15, 0.15, 0.15, 0.4), ncol = 3, byrow = TRUE)

# parameter values
param.value_n <- data.frame(id = NA, shape = NA, scale = NA, kappa = NA) # non_infected 
param.value_i <- data.frame(id = NA, shape = NA, scale = NA, kappa = NA) # infected

# number of individual
n_ind <- 1 

# Initial start point
start <- make_start(c(runif(1, 300000, 360000), runif(1, 1600000, 1700000)), ta_ = 0)

# number of total steps to generate
total_days <- 30
```


## transitional matrix - infection x behavioral states 

prep for the recursive Markov chain manner for getting the infection & behavioral status (each is per day) and 24 steps per day since it is resampled by hourly. 
```{r tmp status generator}
# Modify the code to use a 4x4 transitional probability matrix and track 4 different states daily.

# Define the transitional probability matrix for susceptible, infection, recovered, and dead
# susceptible --> susceptible or infected or natually dead 
# infected --> infected or recovered or infected dead 
# recovered --> only recovered 
# dead --> only dead
trans_m <- matrix(c(
  0.65, 0.3, 0, 0.05,
  0, 0.45, 0.15, 0.4,
  0, 0, 1, 0,
  0, 0, 0, 1
), nrow = 4, ncol = 4, byrow = TRUE)

# Function: ReMC
# Description: This function represents a recursive representation of a Markov chain.
# Parameters:
#   - v: vector representing the initial state of the Markov chain
#   - n: integer representing the number of steps in the Markov chain
# Returns:
#   - A vector representing the state of the Markov chain after n steps

ReMC <- function(v, n) {
  # Base case: if n equals 1, return the initial state v
  if (n == 1) return(v)
  
  # Recursive step: compute the state of the Markov chain after n - 1 steps
  p <- ReMC(v, n - 1)
  
  # Extract the last element of the computed state vector
  u <- tail(p, 1)
  
  # Select the next state based on the transition matrix trans_m
  # Sample from a vector of states with probabilities determined by the transition matrix
  next_state <- sample(1:4, 1, prob = trans_m[u, ])
  
  # Concatenate the computed state vector with the next state
  # This represents the state of the Markov chain after n steps
  c(p, next_state)
}

# Define the initial state value - starting from the susceptible phase  
x0 <- c(state = 1)  # Assuming the initial state is 1 (susceptible)

# Apply the ReMC function to the list of vectors representing the initial parameter values
state_info <- sapply(x0, ReMC, total_days)

# Display the state information
print(state_info)
```


## simulate the paths

The provided code simulates the path of an individual with specified conditions and annotations. It incorporates a transitional probability matrix (tpm) to determine changes in infection status during each step of the simulation. Depending on the current infection status of the individual, the code selects the appropriate probability from the matrix to determine the next state. The simulation iterates through each step, updating the infection status accordingly and simulating the path based on the redistribution kernel.
```{r simulation}
# Create empty lists to store results
ssf_sim <- vector("list", length = total_days)

# Looping for transitions between states
for (ind in 1:n_ind) {
  
  # Model per group
  
  ## model 1 - susceptible
  m_ne <- make_issf_model(coefs = c("sl_" = 0, 
                                    "log(sl_)" = 0, 
                                    "cos(ta_)" = 0),
                          sl = make_gamma_distr(shape = 0.5,
                                                scale = 800),
                          ta = make_vonmises_distr(kappa = 2))
  
  ## model 2 - infected
  m_nr <- make_issf_model(coefs = c("sl_" = 0, 
                                    "log(sl_)" = 0, 
                                    "cos(ta_)" = 0),
                          sl = make_gamma_distr(shape = 0.2,
                                                scale = 300),
                          ta = make_vonmises_distr(kappa = 0.06))
  
  ## model 3 - recovered
  m_ie <- make_issf_model(coefs = c("sl_" = 0, 
                                    "log(sl_)" = 0, 
                                    "cos(ta_)" = 0),
                          sl = make_gamma_distr(shape = 0.6,
                                                scale = 700),
                          ta = make_vonmises_distr(kappa = 0.1))
  
  ## model 4 - dead
  m_ir <- make_issf_model(coefs = c("sl_" = 0, 
                                    "log(sl_)" = 0, 
                                    "cos(ta_)" = 0),
                          sl = make_gamma_distr(shape = 0.1,
                                                scale = 50),
                          ta = make_vonmises_distr(kappa = 0.58))

  # Redistribution kernel per group
  
  ## redistribution kernel - non-infected explore
  rdk_ne <- redistribution_kernel(m_ne, 
                                  start = start, 
                                  map = shrub, 
                                  landscape = "continuous", 
                                  tolerance = 0.02)
  
  ## redistribution kernel - non-infected resting
  rdk_nr <- redistribution_kernel(m_nr, 
                                  start = start, 
                                  map = shrub, 
                                  landscape = "continuous", 
                                  tolerance = 0.02)
  
  ## redistribution kernel - infected exploratory
  rdk_ie <- redistribution_kernel(m_ie, 
                                  start = start, 
                                  map = shrub, 
                                  landscape = "continuous", 
                                  tolerance = 0.02)
  
  ## redistribution kernel - infected resting
  rdk_ir <- redistribution_kernel(m_ir, 
                                  start = start, 
                                  map = shrub, 
                                  landscape = "continuous", 
                                  tolerance = 0.02)

  # Initialize the start position
  start_position <- start
  
  # Loop for each step
  for (i in 1:total_days) {
    # get the infection status of each day 
    state <- state_info[i]
  
    # IF the daily state is noninfected & explore (1), then generate 24 steps by using rdk_n for 1 step-ahead projections
    if (state == 1) {
      ssf_sim[[i]] <- data.frame(id = ind, 
                                 state = state,
                                 positions = list(simulate_path(rdk_ne, n.steps = 24, custom_start = start_position)))
    } else {
      # IF the daily state is noninfected & resting (2), then generate 24 steps by using rdk_i for 1 step-ahead projections
      if (state == 2) {
        ssf_sim[[i]] <- data.frame(id = ind, 
                                    state = state,
                                    positions = list(simulate_path(rdk_nr, n.steps = 24, custom_start = start_position)))
      } else {
        # IF the daily state is infected & explore (3), then generate 24 steps by using rdk_r for 1 step-ahead projections
        if (state == 3) {
          ssf_sim[[i]] <- data.frame(id = ind, 
                                     state = state,
                                     positions = list(simulate_path(rdk_ie, n.steps = 24, custom_start = start_position)))
        } else {
          # IF the daily state is infected & resting (4), then generate 24 steps by using rdk_e for 1 step-ahead projections
          if (state == 4) {
            ssf_sim[[i]] <- data.frame(id = ind, 
                                       state = state,
                                       positions = list(simulate_path(rdk_ir, n.steps = 24, custom_start = start_position)))
          }
        }
      }
    }
    # Update the start position for the next iteration
    start_position <- tail(ssf_sim[[i]]$positions, 1)$position
  }
}

# Combine results into one data frame
ssf_df <- do.call(rbind, ssf_sim)
ssf_df <- ssf_df[!duplicated(ssf_df$positions.x_, ssf_df$positions.y_),]

#positions.t <- seq(as.POSIXct("2024-02-05 20:27:26"), as.POSIXct("2024-02-26 16:27:26"), by="hour")
#ssf_df$positions.t_ <- positions.t

#ssf_df <- ssf_df %>% mutate(postions.t_ = row_number())
ssf_df

# save the data
#saveRDS(ssf_df, "ssf_sim_trial_15days.rds")
#saveRDS(ssf_df, "ssf_sim_trial_15days_same_Rdist.rds")
```

# Visualizations

visualization of the tracks:
```{r trk visualization}
# prep for sff_df
ssf_df$state <- as.factor(ssf_df$state)

# change the labels
state_names <- c('1' = "S", '2' = "I", '3' = "R", '4' = "D")

ggplot()+
  geom_point(data = ssf_df, aes(positions.x_, positions.y_))+
  facet_wrap(~state, 
             nrow = 1,
             labeller = labeller(state = state_names))


# visualize the movement paths across behaviors
p <- ggplot()+
  geom_spatraster(data = shrub, show.legend = FALSE)+
  geom_point(data = ssf_df, aes(positions.x_, positions.y_), size = 0.7)+
  labs(title = "Oryx SIR simulation for 30 days")+
  facet_wrap(~state,
             labeller = labeller(state = state_names),
             nrow = 1)+
  scale_fill_terrain_c()+
  theme_bw()+ 
   coord_sf(
    xlim = range(ssf_df$positions.x_, na.rm = TRUE) + c(-1000, 1000),  # add buffer
    ylim = range(ssf_df$positions.y_, na.rm = TRUE) + c(-1000, 1000)
  )+ 
  theme(axis.text.x=element_text(angle=45, hjust=1))

p
```

# Footer
```{r footer}
sessionInfo()
```