---
title: "7b.real_ui_dat_HMM"
author: "Dennis Kim"
date: "2024-01-22"
output: html_document
---

# Document Preamble 
```{r preamble, message=FALSE, warning=FALSE}
# load libraries
library(knitr)
library(dplyr)
library(readr)
library(data.table)
library(here)
library(stringr)
library(tidyr)
library(purrr)
library(lubridate)
library(scales)

# call environmental data
library(terra)

# analysis
library(momentuHMM)
library(adehabitatLT)

# visualization 
library(ggplot2)

options(width = 150)
```

## 1.1. Format of tracking data

Read in the prep data
```{r read in prep data}
# Location data of the oryx 
whole_df <- readr::read_rds(here::here("data/HMM", "HMM_prep_data.Rdata"))
u_oryx <- read_rds(here::here("data/ssf_data", "u_ssf_oryx.rds"))
i_oryx <- read_rds(here::here("data/ssf_data", "i_ssf_oryx.rds"))

# convert the shrub cover values to percentage 
whole_df <- whole_df %>% mutate(per_shrub = (shrub/100))

# create bern body condition index - if optimal or over = 0, else under = 1
#whole_df <- whole_df %>% mutate(BC = ifelse(c_avg_BC == "under", 1, 0))

# unknown and infected list 
u_i_oryx <- rbind(u_oryx, i_oryx)
u_i_list <- unique(u_i_oryx$ID)

# filter by unknown and infected list 
u_i_prep <- whole_df %>% filter(ID %in% u_i_list)

rm(u_i_oryx, u_i_list, whole_df, i_oryx, u_oryx) # remove the unnecessary dataset

# make sure each individual start with the state from 0 
u_i_prep <- u_i_prep %>% 
  group_by(ID) %>% # grouping by ID
  mutate(row_number = row_number()) %>% # adding a row number for each group 
  mutate(infection = ifelse(row_number <= 24, 0, infection)) %>% # assigning value 0 to the first 10 rows per group
  dplyr::select(- c(row_number, shrub, c_avg_BC)) %>% # un-select the row_number column
  ungroup(ID)

hmm_prep_dat <- data.frame(u_i_prep) # convert it to the dataframe
```

prepare data for HMM format
```{r HMM formats}
# prepare data for HMM (compute step lengths and turning angles)
data_hmm <- momentuHMM::prepData(hmm_prep_dat, type = "UTM")

# check the hmm dataset
head(data_hmm, 10)

# mean of non infected step and turn angle
data_hmm %>% filter(infection == 0) %>% drop_na(.) %>% summarise(., step_mean = mean(step))
data_hmm %>% filter(infection == 0) %>% drop_na(.) %>% summarise(., ta_mean = mean(angle))

# mean of infected step and turn angle
data_hmm %>% filter(infection == 1) %>% drop_na(.) %>% summarise(., step_mean = mean(step))
data_hmm %>% filter(infection == 1) %>% drop_na(.) %>% summarise(., ta_mean = mean(angle))

data_hmm %>% filter(step >= 500)
```

# model fitting - a 4-state HMM

Choose some hypothesised mean step length for each state by looking at the empirical distribution. 
```{r paramter values}
# look at the step lengths distribution
ggplot(data_hmm, aes(x = step))+
  geom_histogram(col = "white", fill = "grey", bins = 20)+
  facet_wrap(~infection, scales = "free")

# look at the infection distribution
ggplot(data_hmm, aes(x = infection))+
  geom_histogram(col = "white", fill = "grey", bins = 2)
```

```{r 4 dependent parameter values}
### Set up model

# number of states 
nbStates <- 4

# label states
stateNames <- c('1', '2', '3', '4')

# Observation distributions (step lengths, turning angles)
dist <- list(step = "gamma", 
             angle = "vm")

# constrain transition probabilities - fix the transition probability matrix to prevent some of the transitions - we set to NA the columns of unconstrained transition probabilities, and we again fix the intercept of the other columns to a large negative number (here -1e6) to set the corresponding transition probabilities to be virtually zero (i.e., impossible transition).
fixbeta <-  matrix(c(NA, -1e6, -1e6,
                     NA, -1e6, NA,
                     -1e6, -1e6, NA,
                     -1e6, -1e6, NA), 
                   nrow = 1, 
                   byrow = TRUE)

# fix initial values 
fixPar <- list(beta = fixbeta)

# Initial parameters
# (step mean for state 1, step mean for state 2, step mean for state N so on... same for step SD s1, step SD s2) and (angle concentration s1, angle concentration s2)
Par0_4s <- list(step = c(400, 50, 100, 50,
                         400, 50, 100, 50), 
                angle = c(0.005, 0.05, 0.002, 0.05)
                )

# Fit a 4-state HMM
hmm <- momentuHMM::fitHMM(data_hmm, 
                          nbStates = nbStates, 
                          fixPar = fixPar,
                          dist = dist, 
                          Par0 = Par0_4s,
                          stateNames = stateNames)
```

model visualization - hmm
```{r hmm model output}
# Print parameter estimates
hmm
```

```{r hmm visualization}
# Plot estimated distributions and state-coloured tracks
plot(hmm, breaks = 20, ask = FALSE)
```

save the plots as a pdf
```{r example, echo=FALSE, results='hide'}
pdf("real_ui_HMM_Oryx.pdf")
plot(hmm, breaks = 20, ask=FALSE)
dev.off()
```

take out the estimates from the fitted model HMM 
```{r model outputs}
# bring out the reconstructed most probable states sequence 
states <- momentuHMM::viterbi(hmm)

# add states into hmm
data_hmm$states <- states

# set up the infection statue from the initial data
data_hmm$infection <- hmm_prep_dat$infection
```

visualization to HMM state sequence over time 
```{r state visualization}
hmm_df <- as.data.frame(data_hmm) # convert it to dataframe 
hmm_df <- hmm_df %>% mutate(julian = yday(time),
                            week = lubridate::week(time)) # create the week column
hmm_df$states <- as.factor(states) # factorize the state
hmm_df$ID <- as.numeric(hmm_df$ID) # convert the ID to numeric
hmm_df <- hmm_df %>% arrange(desc(ID)) # rearrange the ID by desc order
hmm_df_trends <- hmm_df %>% dplyr::select(ID, julian, week, states)

# visualize the trends 
ggplot(hmm_df_trends, aes(x = julian, y = as.factor(ID)))+
       #geom_point(stat='identity', aes(col=states), size = 2)+
       geom_point(stat = 'identity', aes(col=states), size = 1)+
       scale_y_discrete("ID", 
                        labels = as.character(hmm_df$ID), 
                        breaks = hmm_df$ID, 
                        expand = expansion(mult = c(0, .02),add = c(1, 0)))+
       theme_bw()
```

# Footer
```{r footer}
sessionInfo()
```
