---
title: "6.fit_HMM_oryx"
author: "Dennis Kim"
date: "2024-01-22"
output: html_document
---

# Document Preamble 
```{r preamble, message=FALSE, warning=FALSE}
# load libraries
library(knitr)
library(dplyr)
library(readr)
library(data.table)
library(here)
library(stringr)
library(tidyr)
library(purrr)
library(lubridate)
library(scales)

# call environmental data
library(terra)

# analysis
library(momentuHMM)
library(adehabitatLT)

# visualization 
library(ggplot2)

options(width = 150)
```

## 1.1. Format of tracking data

Read in the prep data
```{r read in prep data}
# Location data of the oryx 
whole_df <- readr::read_rds(here::here("data/HMM", "HMM_prep_data.Rdata"))
n_oryx <- read_rds(here::here("data/ssf_data", "n_ssf_oryx.rds"))
u_oryx <- read_rds(here::here("data/ssf_data", "u_ssf_oryx.rds"))
i_oryx <- read_rds(here::here("data/ssf_data", "i_ssf_oryx.rds"))

# convert the shrub cover values to percentage 
whole_df <- whole_df %>% mutate(per_shrub = (shrub/100))

# create bern body condition index - if optimal or over = 0, else under = 1
#whole_df <- whole_df %>% mutate(BC = ifelse(c_avg_BC == "under", 1, 0))

# non-infecetd list 
n_list <- unique(n_oryx$ID)

# unknown and infected list 
u_i_oryx <- rbind(u_oryx, i_oryx)
u_i_list <- unique(u_i_oryx$ID)

# filter by unknown and infected list 
u_i_prep <- whole_df %>% filter(ID %in% u_i_list)
n_prep <- whole_df %>% filter(ID %in% n_list)

rm(u_i_oryx, u_i_list, n_list,  whole_df, i_oryx, u_oryx, n_oryx) # remove the unnecessary dataset

# make sure each individual start with the state from 0 
u_i_prep <- u_i_prep %>% 
  group_by(ID) %>% # grouping by ID
  mutate(row_number = row_number()) %>% # adding a row number for each group 
  mutate(infection = ifelse(row_number <= 24, 0, infection)) %>% # assigning value 0 to the first 10 rows per group
  dplyr::select(- c(row_number, shrub, c_avg_BC)) %>% # un-select the row_number column
  ungroup(ID)

u_i_prep <- data.frame(u_i_prep) # convert it to the dataframe

# set infection status for the non-infected as 0 
n_prep <- n_prep %>% mutate(infection = 0) %>% dplyr::select(- c(shrub, c_avg_BC))
n_prep

# bind two data into hmm prep_dat
hmm_prep_dat <- rbind(n_prep, u_i_prep)
```

prepare data for HMM format
```{r HMM formats}
# prepare data for HMM (compute step lengths and turning angles)
data_hmm <- momentuHMM::prepData(hmm_prep_dat, type = "UTM")

# check the hmm dataset
head(data_hmm, 10)
```

separate the selected population data into individual data
```{r individual data}
# get the list of infected individuals
hmm_prep_dat %>% filter(infection == 0) %>% distinct(ID)

# get the unknown individuals that have more than 700 rows
#hmm_prep_dat %>% nest(data = -ID) %>% mutate(n = map_dbl(data, nrow)) %>% filter(n > 600)

# separate the infected data by  ID 
i_11_hmm <- data_hmm %>% filter(ID == "11")
i_22_hmm <- data_hmm %>% filter(ID == "22")

# seaparate the unknown data by ID 
u_5_hmm <- data_hmm %>% filter(ID == "5")
u_27_hmm <- data_hmm %>% filter(ID == "27")

# non-infected data by ID
n_12_hmm <- data_hmm %>% filter(ID == "12")
n_21_hmm <- data_hmm %>% filter(ID == "21")
```

# model fitting - a 4-state HMM

Choose some hypothesised mean step length for each state by looking at the empirical distribution. 
```{r paramter values}
# look at the step lengths distribution
ggplot(data_hmm, aes(x = step))+
  geom_histogram(col = "white", fill = "grey", bins = 20)+
  facet_wrap(~infection, scales = "free")

# look at the infection distribution
ggplot(data_hmm, aes(x = infection))+
  geom_histogram(col = "white", fill = "grey", bins = 2)
```

# infection movement parameters
get the initial movement parameters 
```{r movement params}
# infected
i_11_hmm %>% 
  group_by(infection) %>% 
  dplyr::summarize(step_mean = mean(step, na.rm = TRUE),
                   ta_mean = mean(angle, na.rm = TRUE))

i_22_hmm %>% 
  group_by(infection) %>% 
  dplyr::summarize(step_mean = mean(step, na.rm = TRUE),
                   ta_mean = mean(angle, na.rm = TRUE))

# unknown 
u_5_hmm %>% 
  group_by(infection) %>% 
  dplyr::summarize(step_mean = mean(step, na.rm = TRUE),
                   ta_mean = mean(angle, na.rm = TRUE))
u_27_hmm %>% 
  group_by(infection) %>% 
  dplyr::summarize(step_mean = mean(step, na.rm = TRUE),
                   ta_mean = mean(angle, na.rm = TRUE))


# non-infected
n_12_hmm %>% 
  group_by(infection) %>% 
  dplyr::summarize(step_mean = mean(step, na.rm = TRUE),
                   ta_mean = mean(angle, na.rm = TRUE))
n_21_hmm %>% 
  group_by(infection) %>% 
  dplyr::summarize(step_mean = mean(step, na.rm = TRUE),
                   ta_mean = mean(angle, na.rm = TRUE))
```

# initital parameters 
```{r initial parameters}
# Initial parameters
# (step mean for state 1, step mean for state 2, step mean for state N so on... same for step SD s1, step SD s2) and (angle concentration s1, angle concentration s2)

# infected
Par0_4s_i11 <- list(step = c(305, 50, 150, 50,
                             305, 50, 150, 50), 
                angle = c(0.06, 0.3, 0.02, 0.3))

Par0_4s_i22 <- list(step = c(300, 50, 150, 50,
                             300, 50, 150, 50), 
                angle = c(0.04, 0.3, 0.02, 0.3))

# unknown
Par0_4s_u5 <- list(step = c(570, 100, 400, 100,
                            570, 100, 400, 100), 
                angle = c(0.07, 0.1, 0.03, 0.1))

Par0_4s_u27 <- list(step = c(1000, 200, 700, 200,
                             1000, 200, 700, 200), 
                angle = c(0.08, 0.1, 0.03, 0.1))

# non-infected
Par0_4s_n12 <- list(step = c(444, 50, 250, 50,
                             444, 50, 250, 50), 
                angle = c(0.03, 0.1, 0.05, 0.1))

Par0_4s_n21 <- list(step = c(495, 50, 350, 50,
                             495, 50, 350, 50), 
                angle = c(0.02, 0.1, 0.05, 0.1))
```


# infected individual fit
```{r infected}
### Set up model

# number of states 
nbStates <- 4

# label states
stateNames <- c('1' = 'NE', '2' = 'NR', '3' = 'IE', '4' = 'IR')

# Observation distributions (step lengths, turning angles)
dist <- list(step = "gamma", 
             angle = "vm")

# constrain transition probabilities - fix the transition probability matrix to prevent some of the transitions - we set to NA the columns of unconstrained transition probabilities, and we again fix the intercept of the other columns to a large negative number (here -1e6) to set the corresponding transition probabilities to be virtually zero (i.e., impossible transition).
fixbeta <-  matrix(c(NA, -1e6, -1e6,
                     NA, -1e6, NA,
                     -1e6, -1e6, NA,
                     -1e6, -1e6, NA), 
                   nrow = 1, 
                   byrow = TRUE)

# fix initial values 
fixPar <- list(beta = fixbeta)
```

# unknown movement models
```{r unknown}
# Fit a 4-state HMM - infected
hmm_11 <- momentuHMM::fitHMM(i_11_hmm, 
                          nbStates = nbStates, 
                          fixPar = fixPar,
                          dist = dist, 
                          Par0 = Par0_4s_i11,
                          stateNames = stateNames)

hmm_22 <- momentuHMM::fitHMM(i_22_hmm, 
                          nbStates = nbStates, 
                          fixPar = fixPar,
                          dist = dist, 
                          Par0 = Par0_4s_i22,
                          stateNames = stateNames)


# Fit a 4-state HMM - unknown
hmm_5 <- momentuHMM::fitHMM(u_5_hmm, 
                          nbStates = nbStates, 
                          fixPar = fixPar,
                          dist = dist, 
                          Par0 = Par0_4s_u5,
                          stateNames = stateNames)

hmm_27 <- momentuHMM::fitHMM(u_27_hmm, 
                          nbStates = nbStates, 
                          fixPar = fixPar,
                          dist = dist, 
                          Par0 = Par0_4s_u27,
                          stateNames = stateNames)


# Fit a 4-state HMM - non-infected
hmm_12 <- momentuHMM::fitHMM(n_12_hmm, 
                          nbStates = nbStates, 
                          fixPar = fixPar,
                          dist = dist, 
                          Par0 = Par0_4s_n12,
                          stateNames = stateNames)

hmm_21 <- momentuHMM::fitHMM(n_21_hmm, 
                          nbStates = nbStates, 
                          fixPar = fixPar,
                          dist = dist, 
                          Par0 = Par0_4s_n21,
                          stateNames = stateNames)
```

model visualization - hmm
```{r hmm model output}
# non-infected
hmm_11
hmm_22

# unknown
hmm_5
hmm_27

# non-infected
hmm_12
hmm_21
```

```{r hmm visualization}
# Plot estimated distributions and state-coloured tracks
# infected
plot(hmm_11, breaks = 25, ask = FALSE)
plot(hmm_22, breaks = 25, ask = FALSE)

# unknown
plot(hmm_5, breaks = 25, ask = FALSE)
plot(hmm_27, breaks = 25, ask = FALSE)

# non-infected
plot(hmm_12, breaks = 25, ask = FALSE)
plot(hmm_21, breaks = 25, ask = FALSE)
```

save the plots as a pdf
```{r example, echo=FALSE, results='hide'}
pdf("real_individual_HMM_Oryx.pdf")
plot(hmm_11, breaks = 25, ask = FALSE)
plot(hmm_22, breaks = 25, ask = FALSE)
plot(hmm_5, breaks = 25, ask = FALSE)
plot(hmm_27, breaks = 25, ask = FALSE)
plot(hmm_12, breaks = 25, ask = FALSE)
plot(hmm_21, breaks = 25, ask = FALSE)
dev.off()
```

# Footer
```{r footer}
sessionInfo()
```