---
title: "6.fit_HMM_oryx"
author: "Dennis Kim"
date: "2024-01-22"
output: html_document
---

# Document Preamble 
```{r preamble, message=FALSE, warning=FALSE}
# load libraries
library(knitr)
library(dplyr)
library(readr)
library(data.table)
library(here)
library(stringr)
library(tidyr)
library(purrr)
library(lubridate)
library(scales)

# call environmental data
library(terra)

# analysis
library(moveHMM)
library(hmmTMB)
library(adehabitatLT)

# visualization 
library(ggplot2)

options(width = 150)
```

## 1.1. Format of tracking data

Read in the prep data
```{r read in prep data}
# Location data of the oryx 
whole_df <- readr::read_rds(here::here("data/HMM", "HMM_prep_data.Rdata"))

# transform ID to factor for random effects
whole_df$ID <- factor(whole_df$ID)

# only select variables of interests 
whole_df <- whole_df %>% dplyr::select(-c(shrub, c_avg_BC))

# look at the data structure
head(whole_df)
```

prepare data for HMM format
```{r HMM formats}
# prepare data for HMM (compute step lengths and turning angles)
data_hmm <- momentuHMM::prepData(whole_df, type = "UTM")

# factor the ID 
data_hmm$ID <- factor(data_hmm$ID)

# check the hmm dataset
head(data_hmm, 10)
```

# model fitting - a 4-state HMM

Choose some hypothesised mean step length for each state by looking at the empirical distribution. 
```{r paramter values}
# look at the step lengths distribution
ggplot(data_hmm, aes(x = step))+
  geom_histogram(col = "white", fill = "grey", bins = 20)+
  facet_wrap(~infection, scales = "free")

# look at the infection distribution
ggplot(data_hmm, aes(x = infection))+
  geom_histogram(col = "white", fill = "grey", bins = 2)
```

# infection movement parameters
get the initial movement parameters 
```{r movement params}
# mean movement params
data_hmm %>% 
  group_by(infection) %>% 
  dplyr::summarize(step_mean = mean(step, na.rm = TRUE),
                   ta_mean = mean(angle, na.rm = TRUE))
```

# Observation model 
```{r obs model}
# Initial parameters
# step mean for state 1, 2, 3, 4
# turn angle for state 1, 2, 3, 4

Par0_4s <- list(step = list(mean = c(500, 100, 300, 100), 
                            sd = c(500, 100, 300, 100)),
                angle = list(mu = c(0, 0, 0, 0), 
                             kappa = c(0.01, 0.15, 0.05, 0.15))
                )

# Observation distributions (step lengths, turning angles)
dist <- list(step = "gamma2", 
             angle = "vm")

# create Observation object 
obs <- Observation$new(data = data_hmm,
                       dists = dist,
                       n_states = 4,
                       par = Par0_4s)
```

# Hidden state model 

create a matrix to specify the structure of the model, where each element is the formula for the corresponding transition probability. 
```{r hidden state model}
# Model formulas 
f <- "~ s(ID, bs = 're')"

# the diagnoal elements are used as reference for each row, so their formulas are set to ".". 
tpm_structure <- matrix(c(".",    f, "~1", "~1",
                            f,  ".", "~1",    f,
                         "~1", "~1",  ".",    f,
                         "~1", "~1",    f,  "."),
                        nrow = 4,
                        byrow = TRUE)


# Initial transition probabilites
#tpm0 <-  matrix(c(0.7, 0.3, 0, 0,
#                  0.12, 0.48, 0, 0.4,
#                  0, 0, 0.2, 0.8,
#                  0, 0, 0.1, 0.9),
#                nrow = 4, 
#                byrow = TRUE)

# create MarkovChain object - use the option initial_state = "stationary" to fix the initial distribution of the state process to the stationary distirbution for each time series. 
hid <- MarkovChain$new(n_states = 4,
                       formula = tpm_structure,
                       data = data_hmm,
                       initial_state = "stationary")

hid
```

list fixed parameters 
```{r fixed paramters}
# list of fixed parameters 
fixpar <- list(obs = c("angle.mu.state1.(Intercept)" = NA,
                       "angle.mu.state2.(Intercept)" = NA,
                       "angle.mu.state3.(Intercept)" = NA,
                       "angle.mu.state4.(Intercept)" = NA),
               hid = c("S1>S3.(Intercept)" = 0,
                       "S1>S4.(Intercept)" = 0,
                       "S2>S3.(Intercept)" = 0,
                       "S3>S1.(Intercept)" = 0,
                       "S3>S2.(Intercept)" = 0,
                       "S4>S1.(Intercept)" = 0,
                       "S4>S2.(Intercept)" = 0))

# create HMM object 
hmm <- HMM$new(obs = obs, hid = hid, fixpar = fixpar)
```

# fit the HMM model 
```{r model fit}
# Fit a 4-state HMM
hmm$fit(silent = TRUE)
```

model visualization - hmm
```{r hmm model output}
hmm
```

# interpreting the state
```{r hmm visualization}
hmm$plot_dist("step")+
  coord_cartesian(ylim = c(0, 0.0005))+
  theme(legend.position = c(0.8, 0.8))+
  labs(x = "step (m)")
```

# inter-individual heterogeneity 
```{r inter}
hmm$plot(what = "delta", var = "ID")
```

save the plots as a pdf
```{r example, echo=FALSE, results='hide'}
pdf("real_individual_HMM_Oryx.pdf")
plot(hmm_11, breaks = 25, ask = FALSE)
dev.off()
```

# Footer
```{r footer}
sessionInfo()
```