---
title: "6.fit_HMM_oryx"
author: "Dennis Kim"
date: "2024-01-22"
output: html_document
---

# Document Preamble 
```{r preamble, message=FALSE, warning=FALSE}
# load libraries
library(knitr)
library(dplyr)
library(readr)
library(data.table)
library(here)
library(stringr)
library(tidyr)
library(purrr)
library(lubridate)
library(scales)

# call environmental data
library(terra)

# analysis
library(momentuHMM)
library(adehabitatLT)

# visualization 
library(ggplot2)

options(width = 150)
```

## 1.1. Format of tracking data

Read in the prep data
```{r read in prep data}
# Location data of the oryx S
sim_df <- readr::read_rds(here::here("data/simulation_data", "ssf_sim_trk1_60days.rds"))

# align with the time data
sim_df <- head(sim_df, -3)

# create time wiht intervals of 1 hour
sim_df$positions.t_ <- seq(as.POSIXct("2022-02-01"), as.POSIXct("2022-04-01"), by = "hour")

# change the column names 
colnames(sim_df) <- c("ID", "state", "x", "y", "time", "dt")

# data summary
summary(sim_df)
```

prepare data for HMM format
```{r HMM formats}
# prepare data for HMM (comput step lengths and turning angles)
data_hmm <- momentuHMM::prepData(sim_df, type = "UTM", covNames = "state")

# check the hmm dataset
head(data_hmm, 10)

# non-infected exploring: mean of non infected step and turn angle
data_hmm %>% filter(state == 1) %>% drop_na(.) %>% summarise(., step_mean = mean(step))
data_hmm %>% filter(state == 1) %>% drop_na(.) %>% summarise(., ta_mean = mean(angle))

# non-infected resting: mean of infected step and turn angle
data_hmm %>% filter(state == 2) %>% drop_na(.) %>% summarise(., step_mean = mean(step))
data_hmm %>% filter(state == 2) %>% drop_na(.) %>% summarise(., ta_mean = mean(angle))

# infected exploring: mean of non infected step and turn angle
data_hmm %>% filter(state == 3) %>% drop_na(.) %>% summarise(., step_mean = mean(step))
data_hmm %>% filter(state == 3) %>% drop_na(.) %>% summarise(., ta_mean = mean(angle))

# infected resting: mean of infected step and turn angle
data_hmm %>% filter(state == 4) %>% drop_na(.) %>% summarise(., step_mean = mean(step))
data_hmm %>% filter(state == 4) %>% drop_na(.) %>% summarise(., ta_mean = mean(angle))
```

# model fitting - a 2-state HMM

Choose some hypothesised mean step length for each state by looking at the empirical distribution. 
```{r paramter values}
# look at the step lengths distribution
ggplot(data_hmm, aes(x = step))+
  geom_histogram(col = "white", fill = "grey", bins = 20)+
  facet_wrap(~state, scales = "free")
```

```{r 4 dependent parameter values}
### Set up model

# number of states 
nbStates <- 5

# label states
stateNames <- c('1' ="NE", '2' = "NR", '3' = "IE", '4' = "IR", '5' = "D")

# Observation distributions (step lengths, turning angles)
dist <- list(step = "gamma", 
             angle = "vm")

# constrain transition probabilities - fix the transition probability matrix to prevent some of the transitions - we set to NA the columns of unconstrained transition probabilities, and we again fix the intercept of the other columns to a large negative number (here -1e6) to set the corresponding transition probabilities to be virtually zero (i.e., impossible transition).
#1->2 1->3 1->4 1->5
#2->1 2->3 2->4 2->5
#3->1 3->2 3->4 3->5
#4->1 4->2 4->3 4->5
#5->1 5->2 5->3 5->4

fixbeta <-  matrix(c(NA, -1e6, -1e6, -1e6,
                     NA, -1e6, NA, -1e6,
                     -1e6, -1e6, NA, NA,
                     -1e6, -1e6, NA, NA,
                     -1e6, -1e6, -1e6, -1e6), 
                   nrow = 1, 
                   byrow = TRUE)


# fix initial values 
fixPar <- list(beta = fixbeta)

# Initial parameters
# (step mean for state 1, step mean for state 2, step mean for state N so on... same for step SD s1, step SD s2) and (angle concentration s1, angle concentration s2)
Par0_4s <- list(step = c(16000, 7500, 10000, 650, 1,
                         16000, 7500, 10000, 650, 1), 
                angle = c(0.01, 0.1, 0.05, 0.1, 0.1)
                )

# Fit a 4-state HMM
hmm <- momentuHMM::fitHMM(data_hmm, 
                           nbStates = nbStates, 
                           dist = dist, 
                           Par0 = Par0_4s,
                           fixPar = fixPar,
                           stateNames = stateNames)
```

model visualization - hmm1
```{r hmm model output}
# Print parameter estimates
hmm
```

```{r hmm visualization}
# Plot estimated distributions and state-coloured tracks
plot(hmm, breaks = 25, ask = FALSE)
```

save the plots as a pdf
```{r example, echo=FALSE, results='hide'}
pdf("simulation_HMM_Oryx.pdf")
plot(hmm, breaks = 25, ask=FALSE)
dev.off()
```

take out the estimates from the fitted model HMM 
```{r model outputs}
# bring out the reconstructed most probable states sequence 
states <- momentuHMM::viterbi(hmm)

# add states into hmm
data_hmm$states <- states

data_hmm$states <- factor(states, levels = c("1", "2", "3", "4", "5"))

# save the rds file for the state sequences 
saveRDS(data_hmm, "simulation_hmm_sequence.rds")
```

visualization to HMM state sequence over time 
```{r state visualization}
hmm_df <- as.data.frame(data_hmm1) # convert it to dataframe 
hmm_df <- hmm_df %>% mutate(julian = yday(time),
                            week = lubridate::week(time)) # create the week column
hmm_df$states <- as.factor(states) # factorize the state
hmm_df$ID <- as.numeric(hmm_df$ID) # convert the ID to numeric
hmm_df <- hmm_df %>% arrange(desc(ID)) # rearrange the ID by desc order
hmm_df_trends <- hmm_df %>% dplyr::select(ID, julian, week, states)

# visualize the trends 
ggplot(hmm_df_trends, aes(x = julian, y = as.factor(ID)))+
       #geom_point(stat='identity', aes(col=states), size = 2)+
       geom_line(stat = 'identity', aes(col=states), size = 1)+
       scale_y_discrete("ID", 
                        labels = as.character(hmm_df$ID), 
                        breaks = hmm_df$ID, 
                        expand = expansion(mult = c(0, .02),add = c(1, 0)))+
       theme_bw()
```

# Footer
```{r footer}
sessionInfo()
```
